---
title: "Impact of Transmission Type on Fuel Efficiency in MTCARS Data Set"
author: "Allen Hammock"
date: "June 13, 2015"
output: html_document
---
```{r, echo = FALSE}
# setup
library(ggplot2)
library(car)
data(mtcars)

figures.echo <- FALSE
```

# Executive Summary
```{r, echo = FALSE}
mtcars$TransType <- ifelse(mtcars$am == 1, "Auto", "Manual")
fit1 <- lm(mpg ~ TransType - 1, mtcars)
fit2 <- lm(mpg ~ wt + TransType - 1, mtcars)
```

At __Motor Trend__, a magazine about the automobile industry, we are interested 
in exploring the relationship between a transmission types and fuel efficiency 
measured in miles per gallon (MPG). We performed a study on `r nrow(mtcars)` 
different models and found, as a group, that the motorcars with Automatic 
transmissions had an overall better mean fuel efficiency of 
`r round(fit1$coef[c("TransTypeAuto")], 1)` MPG versus vehicles with Manual 
transmissions which had an average `r round(fit1$coef[c("TransTypeManual")], 1)` 
MPG.  

However, when we look at cars in different weight classes, we see that the 
transmission type has a much lower influence on fuel efficiency.  The 
coeffecients on our linear model show, that Manual and Automatic transmissions 
have approximately the same influence on fuel efficiency (MPG) if we hold the 
weight constant.

## MPG ~ TRANSMISSION TYPE (NO INTERCEPT)

From our initial fit, we can see the expected value for MPG is just the 
coefficient, or the mean average of the fuel efficiency ratings for each 
transmission type.

```{r, echo = FALSE, message = FALSE}
fit1$call
round(fit1$coef, 2)
#round(confint(fit1), 2)
```

## MPG ~ WEIGHT + TRANSMISSION TYPE (NO INTERCEPT)

When we adjust for the weight of the vehicle in addition to the transmission
type, we see the difference between Manual and Automatic diminish dramatically.

```{r, echo = FALSE}
fit2$call
round(fit2$coef, 2)
#round(confint(fit2), 2)
```

In our analysis below you will see that regression diagnostics support this
conclusion.  We also examine the leverage that other factors such as the 
number of cylinders have on the overall fuel efficiency outcome.

# Regression Diagnostics

## Normality of Residuals

First, we check to see if the residual error in our fitted model are 
normally distributed (__Figure 1__).  Our assumptions that the errors are
roughly `iid` $N(0, \sigma^2)$ seem to hold.

## Residual Fit


## Leverage

## Conclusions


```{r}
```

# Appendix

## Figure 1 -- Normal QQ

```{r, message = FALSE, echo = figures.echo}
fit2.resid <- residuals(fit2)
fit2.yhat <- predict(fit2)
p2 <- ggplot(data = as.data.frame(qqnorm(fit2.resid , plot=F)), 
             mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Normal Q-Q") +
  xlab("Theretical Quantiles") +
  ylab("Standardized Residuals")
p2
```

## Figure 2 - Fitted vs Residuals

```{r, message = FALSE, echo = figures.echo}
df <- data.frame(x = fit2.yhat, y = fit2.resid)
p1 <- ggplot(df, aes(x = fit2.yhat, y = fit2.resid)) +
  geom_point() +
  ggtitle("Residuals vs Fitted") +
  xlab("Fitted Values") +
  ylab("Residuals") + 
  geom_hline(yintercept = 0) + 
  geom_smooth()
p1
```

## Figure 3 - Leverage

```{r, message = FALSE, echo = figures.echo}
fit2.sigma <- sqrt(deviance(fit2)/df.residual(fit2))
fit2.rsquare <- fit2.resid/fit2.sigma 
fit2.sqrt.rsquare <- sqrt(abs(fit2.rsquare))

fit2.hii <- lm.influence(fit2, do.coef = FALSE)$hat
df <- data.frame(x = fit2.hii, y = fit2.rsquare)
p3 <- ggplot(df, aes(x = fit2.hii, y = fit2.rsquare)) +
  geom_point() +
  ggtitle("Residuals vs Leverage") +
  xlab("Cook's Distance") +
  ylab("Standardized Residuals") +
  geom_hline(yintercept = 0) +
  geom_smooth() +
  theme(legend.position = "none")
p3
```
