---
title: "Impact of Transmission Type on Fuel Efficiency in MTCARS Data Set"
author: "Allen Hammock"
date: "June 13, 2015"
output: word_document
---
<style type="text/css">
@media print {
  .pagebreak { 
    page-break-before: always;
  }
}
</style>
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# libraries
library(gridExtra)
library(ggplot2)
library(knitr)
library(car)
data(mtcars)

# fitted models
mtcars$TransType <- ifelse(mtcars$am == 1, "Auto", "Manual")
mtcars$Cylinders <- as.factor(mtcars$cyl)
fit1 <- lm(mpg ~ TransType - 1, mtcars)
fit2 <- lm(mpg ~ wt + TransType - 1, mtcars)
fit3 <- lm(mpg ~ wt + Cylinders + TransType - 1, mtcars)

# supress code?
figures.echo <- FALSE
```

# Executive Summary

_To view the code for this report, check out the 
[repository on Github](https://github.com/brainvat/RegressionModelsProject/blob/master/report.Rmd)
<a href="#ref1">[1]</a>._

At _Motor Trend_, a magazine about the automobile industry, we are interested 
in exploring the relationship between a transmission types and fuel efficiency 
measured in miles per gallon (MPG). We performed a study on `r nrow(mtcars)` 
different models and found, as a group, that the motorcars with Automatic 
transmissions had an overall better mean fuel efficiency of 
`r round(fit1$coef[c("TransTypeAuto")], 1)` MPG versus vehicles with Manual 
transmissions which had an average `r round(fit1$coef[c("TransTypeManual")], 1)` 
MPG.  

However, when we look at _cars in different weight classes_ and consider
how many cylinders each vehicle has, we see that the transmission type has a 
much lower influence on fuel efficiency than these other factors.  As the 
coeffecients on our linear model show, Manual and Automatic transmissions 
have approximately the same influence (less than
`r round(fit2$coef[c("TransTypeManual")] - fit2$coef[c("TransTypeAuto")], 2)`
MPG difference) on fuel efficiency (MPG) if we hold the weight (wt
coefficient) constant.

This is fairly intuitive, the fact that the heavier, "gas guzzling" cars with
bigger engines tend to have a lower MPG suggests that the efficiences of
automatic transmissions may be negligible compared to these other parameters.

__First Model: lm(mpg ~ TransType - 1, mtcars)__

From our initial fit 
([the code](https://github.com/brainvat/RegressionModelsProject/blob/master/report.Rmd)
<a href="#ref1">[1]</a> is hidden to save space), the expected value for MPG is 
just the coefficient, or the mean average of the fuel efficiency ratings for 
each transmission type.  We have removed the intercept in all of our models.

```{r, echo = FALSE, message = FALSE}
summary(fit1)$coefficients
```

__Second Model: lm(mpg ~ wt + TransType - 1, mtcars)__

Now we adjust the outcome for the influence of both weight and transmission 
type.  Here we see the difference between Manual and Automatic diminish 
dramatically. 

```{r, echo = FALSE}
summary(fit2)$coefficients
```

The t-test is significant and our confidence intervals for Manual vs. Automatic
transmissions are very similar. The distribution of our residuals 
are normally distributed (__<a href="#figure1">Figure 1</a>__) for all three
models so our assumptions that the errors are roughly iid $N(0, \sigma^2)$ seem 
to hold.

```{r, echo = FALSE}
round(confint(fit2), 2)
```

__Third Model: lm(mpg ~ wt + Cylinders + TransType - 1, mtcars)__

Finally we update the model to adjust for the total number of cylinders.  As the 
number of cylinders increases, the MPG decreases.  We see that having a
Manual transmission accounts for a very small part of the variance now and
that within the confidence interval the influence could be positive or negative
since it is so close to zero.

In  __<a href="#figure2">Figure 2</a>__ we see a better dispersion of the
residuals vs. the fitted values as we add these additional parameters suggesting
that we have accounted for much of the systematic variance in our model.

We also did some digging for outliers and we found that the __Toyota Corolla__
and __Fiat 128__ are interesting examples to look at, as 
__<a href="#figure3">Figure 3</a>__ suggests.  

```{r, echo = FALSE}
summary(fit3)$coefficients
round(confint(fit3), 2)
```

<div class="pagebreak"></div>
# Appendix

<a name="figure1" />

__Figure 1 -- Normal QQ__

```{r, message = FALSE, warning = FALSE, echo = figures.echo, fig.width = 4, fig.height = 4}
fit1.resid <- residuals(fit1)
fit1.yhat <- predict(fit1)
p1a <- ggplot(data = as.data.frame(qqnorm(fit1.resid , plot=F)), 
             mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("lm(mpg ~ Trans - 1)") +
  xlab("Theretical Quantiles") +
  ylab("Standardized Residuals")

fit2.resid <- residuals(fit2)
fit2.yhat <- predict(fit2)
p1b <- ggplot(data = as.data.frame(qqnorm(fit2.resid , plot=F)), 
             mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("lm(mpg ~ wt + Trans - 1)") +
  xlab("Theretical Quantiles") +
  ylab("Standardized Residuals")

fit3.resid <- residuals(fit3)
fit3.yhat <- predict(fit3)
p1c <- ggplot(data = as.data.frame(qqnorm(fit3.resid , plot=F)), 
             mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("lm(mpg ~ wt + Cyl + Trans - 1)") +
  xlab("Theretical Quantiles") +
  ylab("Standardized Residuals")

p1a
p1b
p1c
```

<a name="figure2" />

__Figure 2 - Fitted vs Residuals__

```{r, message = FALSE, warning = FALSE, echo = figures.echo, fig.width = 4, fig.height = 4}
dfa <- data.frame(x = fit1.yhat, y = fit1.resid)
p2a <- ggplot(dfa, aes(x = fit1.yhat, y = fit1.resid)) +
  geom_point() +
  ggtitle("lm(mpg ~ Trans - 1)") +
  xlab("Fitted Values") +
  ylab("Residuals") + 
  geom_hline(yintercept = 0) + 
  geom_smooth()

dfb <- data.frame(x = fit2.yhat, y = fit2.resid)
p2b <- ggplot(dfb, aes(x = fit2.yhat, y = fit2.resid)) +
  geom_point() +
  ggtitle("lm(mpg ~ wt + Trans - 1)") +
  xlab("Fitted Values") +
  ylab("Residuals") + 
  geom_hline(yintercept = 0) + 
  geom_smooth()

dfc <- data.frame(x = fit3.yhat, y = fit3.resid)
p2c <- ggplot(dfc, aes(x = fit3.yhat, y = fit3.resid)) +
  geom_point() +
  ggtitle("lm(mpg ~ wt + Cyl + Trans - 1)") +
  xlab("Fitted Values") +
  ylab("Residuals") + 
  geom_hline(yintercept = 0) + 
  geom_smooth()

p2a
p2b
p2c
```

<a name="figure3" />

__Figure 3 - Outliers__

```{r, message = FALSE, warning = FALSE, echo = figures.echo, fig.width = 8, fig.height = 4}
# look for outliers
fit1.ot <- outlierTest(fit1)
fit2.ot <- outlierTest(fit2)
fit3.ot <- outlierTest(fit3)

# pull out the vehicles we found
cars.ot <- unique(sapply(seq(1,3), function(x) names(get(paste0("fit", x, ".ot"))$rstudent)))

# examine predictions for each model on the outliers
mtcars[cars.ot, c("mpg", "wt", "Cylinders", "TransType")]
ot.predictions <- sapply(seq(1, 3), 
                           function(x) {
                             predict(get(paste0("fit", x)), 
                                     mtcars[cars.ot, c("mpg", "wt", "Cylinders", "TransType")])})
cars.observed <- mtcars[cars.ot, c("mpg")]
ot.pct <- matrix(paste0(round(100*(ot.predictions - cars.observed)/cars.observed, 2), "%"), nrow = 2)
rownames(ot.pct) <- cars.ot
colnames(ot.pct) <- c("mpg ~ Trans", "mpg ~ Trans", "mpg ~ Cyl + Trans")
kable(ot.pct)
```

__References__

<a name="ref1" />
&lt;1&gt; https://github.com/brainvat/RegressionModelsProject/blob/master/report.Rmd
<a name="ref2" />

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# inspect results later
save.image("report.RData")
```